Differential expression analysis of reprocessed AMPAD RNASeq data (All 7 brain regions with CQN normlaisation and BIC based model selection with study regressed apriori)

### 2017-07-27 code/Rmd/ALL_REPROCESSED_v2.Rmd
Added markdown to process all rnaseq data from three brain banck with study regressed prior to model selection

---
title: "Differential expression analysis of reprocessed AMPAD RNASeq data (All 7 brain regions with CQN normlaisation and BIC based model selection)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

@@

knit2synapse::knitToFolderEntity(file = "./ALL_REPROCESSED.Rmd",
                                 parentId = "syn8672415",
                                 entityName = 'ALL Reprocessed RNASeq (Normalised)')
```

@@

thisFileName <- 'ALL_REPROCESSED.Rmd'

@@

# Intronic Rate
p[[4]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_INTRONIC_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Intronic Fractions') + my.theme

# IntergenicRate
p[[5]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_INTERGENIC_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Intergenic Fractions') + my.theme

# Ribosomal Fractions
p[[6]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_RIBOSOMAL_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Ribosomal Fractions') + my.theme

# Reads Aligned
p[[7]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_PF_READS_ALIGNED, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[7]] = p[[7]] + ggtitle('Fraction Reads Aligned') + my.theme

# Coding Bases
p[[8]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_CODING_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[8]] = p[[8]] + ggtitle('Fraction Coding Bases') + my.theme

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=BrainRegion, shape=Study, size=RIN))
p <- p + theme_bw() + theme(legend.position="right")
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("BrainRegion", "Gender", "Diagnosis", "Study")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

### Model Identification
```{r fxn}
# Function to fit a mixed effect model and return BIC
fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){

@@

  # Estimate voom weights
  voomWeights = limma::voom(cnt[rownames(lcpm),colnames(lcpm)], design = designMat, plot=F)$weights

@@

# Function to select covariates in a step-wise regression fashion
selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){

  varsSelected = c()
  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
  while(length(covarsToSelect) != 0){
    allModelBIC = foreach(i = 1:length(covarsToSelect),
                          .combine = c,
                          .packages = pkgs,
                          .export = c('fitMixedEffectModel')) %dopar% {

                            # Fit a mixed effect linear model and get all BIC
                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
                                                           c('individualIdentifier'), 
                                                           lcpm, 
                                                           cnt, 
                                                           covars) %>%
                            plyr::rename(c('BIC' = covarsToSelect[i]))

@@

    # Select variable
    tmp = nChange %>%
      dplyr::filter(nchange > 1) %>%
      dplyr::top_n(1, nchange)

@@

  return(varsSelected)
}
```
#### Phase I (clinical, ancestry and sample specific technical variables)
```{r phase1}
# Fit a base model with the following covariates
adjust.covars = c('BrainRegion.Diagnosis')
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   c('individualIdentifier'), 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase1.covars = c('Study', 'Gender', 'PMI', 'AOD', 'RIN', 'RIN2')
vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

@@

#### Phase II (batch effects)
```{r phase2}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   c('individualIdentifier'), 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase2.covars = c('Batch')
vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

@@

#### Phase III (RNASeq alignment specific covariates)
```{r phase3}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   c('individualIdentifier'), 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase3.covars = c("PCT_PF_READS_ALIGNED", "PCT_CODING_BASES", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

>>>>

---
title: "Differential expression analysis of reprocessed AMPAD RNASeq data (All 7 brain regions with CQN normlaisation and BIC based model selection with study regressed apriori)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

@@

knit2synapse::knitToFolderEntity(file = "./ALL_REPROCESSED_v2.Rmd",
                                 parentId = "syn8672415",
                                 entityName = 'ALL Reprocessed RNASeq (Normalised) - study effects removed')
```

@@

thisFileName <- 'ALL_REPROCESSED_v2.Rmd'

@@

# Intronic Rate
p[[4]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_INTRONIC_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Intronic Fractions') + my.theme + ylab('Fraction')

# IntergenicRate
p[[5]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_INTERGENIC_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Intergenic Fractions') + my.theme + ylab('Fraction')

# Ribosomal Fractions
p[[6]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_RIBOSOMAL_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Ribosomal Fractions') + my.theme + ylab('Fraction')

# Reads Aligned
p[[7]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_PF_READS_ALIGNED, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[7]] = p[[7]] + ggtitle('Fraction Reads Aligned') + my.theme + ylab('Fraction')

# Coding Bases
p[[8]] = ggplot(COVARIATES, aes(x = BrainRegion, y = PCT_CODING_BASES, color = Study, fill = Diagnosis)) + geom_boxplot()
p[[8]] = p[[8]] + ggtitle('Fraction Coding Bases') + my.theme + ylab('Fraction')

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=BrainRegion, shape=Study, size=RIN))
p <- p + theme_bw() + theme(legend.position="top") + xlab(paste0('PC1(',round(100*PC$sdev[1]^2/sum(PC$sdev^2)),'%)')) 
p <- p + ylab(paste0('PC2(',round(100*PC$sdev[2]^2/sum(PC$sdev^2)),'%)')) 
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("BrainRegion", "Gender", "Diagnosis", "Study")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

### Model Identification
```{r fxn}
# Function to fit a mixed effect model and return BIC
fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, voomWeights, covars){

@@

@@

# Function to select covariates in a step-wise regression fashion
selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, voomWeight, baseModelBIC){

  varsSelected = c()
  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
  while(length(covarsToSelect) != 0){
    allModelBIC = foreach(i = 1:length(covarsToSelect),
                          .combine = c,
                          .packages = pkgs,
                          .export = c('fitMixedEffectModel','voomWeight')) %dopar% {

                            # Fit a mixed effect linear model and get all BIC
                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
                                                           c('individualIdentifier'), 
                                                           lcpm, 
                                                           voomWeight, 
                                                           covars) %>%
                            plyr::rename(c('BIC' = covarsToSelect[i]))

@@

    # Select variable
    tmp = nChange %>%
      dplyr::filter(nchange > 5) %>%
      dplyr::top_n(1, nchange)

@@

  return(varsSelected)
}
```
#### Regress out Study 
```{r regress.study}
# Get design matrix
DESIGN.MAT = getDesignMatrix(COVARIATES[colnames(NEW.COUNTS), 'Study', drop=F], Intercept = F)
DESIGN.MAT = DESIGN.MAT$design[,linColumnFinder(DESIGN.MAT$design)$indepCols]

# Estimate voom weights
VOOM.WEIGHTS = limma::voom(NEW.COUNTS, design = DESIGN.MAT, plot=F)

# Fit a linear model
FIT = lmFit(VOOM.WEIGHTS)

# Calculate residuals
RESIDUAL.EXPR = residuals.MArrayLM(FIT, VOOM.WEIGHTS$E)

# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.EXPR, scale.=T, center = T)

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=BrainRegion, shape=Study, size=RIN))
p <- p + theme_bw() + theme(legend.position="right")
p
```
#### Phase I (clinical, ancestry and sample specific technical variables)
```{r phase1}
# Fit a base model with the following covariates
adjust.covars = c('BrainRegion.Diagnosis')
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   c('individualIdentifier'), 
                                   RESIDUAL.EXPR, 
                                   VOOM.WEIGHTS$weights, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase1.covars = c('Gender', 'PMI', 'AOD', 'RIN', 'RIN2')
vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     VOOM.WEIGHTS$weights, baseModelBIC)

@@

#### Phase II (batch effects)
```{r phase2}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   c('individualIdentifier'), 
                                   RESIDUAL.EXPR, 
                                   VOOM.WEIGHTS$weights, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase2.covars = c('Batch')
vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     VOOM.WEIGHTS$weights, baseModelBIC)

@@

#### Phase III (RNASeq alignment specific covariates)
```{r phase3}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   c('individualIdentifier'), 
                                   RESIDUAL.EXPR, 
                                   VOOM.WEIGHTS$weights, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase3.covars = c("PCT_PF_READS_ALIGNED", "PCT_CODING_BASES", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     VOOM.WEIGHTS$weights, baseModelBIC)
